---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


# Loading and preprocessing the data


```{r}
    ## read the data
    rawData <- read.csv("activity.csv")
    ## create a list of days from the data set
    days <- as.character(unique(rawData$date))
```


# What is mean total number of steps taken per day?
```{r}
   ###Initialize the list of steps with the first day.
    firstDay <- days[1]
    stepsList <- sum(subset(rawData$steps, rawData$date == firstDay, na.rm = T)) 

        ###Add the remaining steps per day to the list
    for( i in 2:length(days)) {
        day <- days[i]
        stepsList <- c(stepsList, sum(subset(rawData$steps, rawData$date == day), na.rm = T) )
    
    }
```    
###A histogram of steps per day 
```{r}
    hist(stepsList, xlab = "Step per day")
```

###The mean of steps per day in the period
```{r}
    print(mean(stepsList, na.rm = T) ) 
```
###The median of steps per day in the period
```{r}
    print(median(stepsList, na.rm = T) )
```


# What is the average daily activity pattern?

```{r}
    ###Initialize the list of mean steps in the first interval.
    intervals <- unique(as.character(rawData$interval))
    firstInterval <- intervals[1]
    meanStepsPerIntervalList <- mean(subset(rawData$steps, rawData$interval == firstInterval), na.rm = T) 

    ###Add the remaining means of steps per interval to the list.
    for( i in 2:length(intervals)) {
        interval <- intervals[i]
        meanStepsPerIntervalList <- c(meanStepsPerIntervalList, mean(subset(rawData$steps, rawData$interval == interval) , na.rm = T ) )
        
    }
    
```
###Mean number of steps per interval
```{r}
    plot(intervals, meanStepsPerIntervalList, type="l")
```

###Max number of steps per interval on average
``` {r}
    IntervalMeanTable <- (cbind (intervals, meanStepsPerIntervalList))
    maxMean <- max(as.numeric(IntervalMeanTable[,2]))
    maxInterval <- subset(IntervalMeanTable, as.numeric(IntervalMeanTable[,2])==maxMean)
    print(maxInterval[1])    
```
# Imputing missing values
### Total number of NA rows
``` {r}
    print( sum( is.na( rawData ) ) )
```
### Impute missing values with the mean of steps during the given interval
``` {r}
    # Add the first imputed value to the mean by interval table. 
    firstImpute <- as.numeric( IntervalMeanTable[1,2] )
    impData <- rawData[1,]
    impData[1,1] <- firstImpute
    # Add remaining imputed values if # of steps are NA in the given interval
    for ( i in 2:dim( rawData )[1]  ) {
        impData[i,] <- rawData[i,]
        if(is.na( rawData[i,1]  ) )  {
            interval <- rawData[i,3]
            nextImputeInterval <- subset(IntervalMeanTable, as.numeric( IntervalMeanTable[,1])==interval)
            impData[i,1] <- nextImputeInterval[2]
        }
    }
```
### Histogram of steps per day with imputed values included

```{r}
    ###Initialize the list of steps with the first day.
    firstDay <- days[1]
    stepsList <- sum(subset(as.numeric(impData$steps), impData$date == firstDay, na.rm = T)) 

    ###Add the remaining steps per day to the list.
    for( i in 2:length(days)) {
        day <- days[i]
        stepsList <- c(stepsList, sum(subset(as.numeric(impData$steps), impData$date == day), na.rm = T) )
    
    }

    hist(stepsList, xlab = "Steps per day")
```

###The mean of steps per day in the period
```{r}
    print(mean(stepsList, na.rm = T) ) 
```
###The median of steps per day in the period
```{r}
    print(median(stepsList, na.rm = T) )
```

### Adding the imputed values increases the number of days with steps between 10,000 and 15,000 a great deal.  It also reduces the number of days with steps from 0 to 5,000 down to 5 instances.  The mean and median numbers of steps per day do not change significantly.

# Are there differences in activity patterns between weekdays and weekends?
###Add logical columns for week ends and week days
``` {r}
    WeekDays <- NULL
    WeekEnds <- NULL
    ##Add TRUE to the WeekEnds list if the given day is Saturday or Sunday.  Also add
    ## FALSE to the WeekDays list in this case.  Do the opposite if the given day is 
    ## Monday through Friday
    for ( i in 1:dim(impData)[1] ) {
        if( weekdays( as.Date( impData[i,2] ) ) == "Saturday" || weekdays( as.Date( impData[i,2] ) ) == "Sunday" ) {
            WeekEnds <- c(WeekEnds, TRUE)
            WeekDays <- c(WeekDays, FALSE)
        }
        else {
            WeekEnds <- c(WeekEnds, FALSE)
            WeekDays <- c(WeekDays, TRUE)
        }
    }
    #Add WeekDays and WeekEnds logical lists to the imputed data frame
    impData <- cbind( impData, WeekDays )
    impData <- cbind( impData, WeekEnds )
```
###Steps on the Week Ends

